%%%\begin{comment}
\begin{thebibliography}{GMS94A}
\small

%%% keep
\bibitem[AHK90]{IMPATIENT}
  Paul W. Abrahams, Kathryn Hargreaves and Karl Berry.
  \newblock \emph{TeX for the Impatient}.
  \newblock Addison-Wesley, 1990.
  \newblock (Available at
             \url{ftp://tug.org/tex/impatient})

\begin{comment}
\bibitem[Ado01]{ADOBEBOOK}
  \emph{How to Create Adobe PDF eBooks}.
  \newblock Adobe Systems Inc.,
  \newblock 2001.
  \newblock (Available from 
             \url{http://www.adobe.com/epaper/tips/acr5ebook/pdfs/eBook.pdf})
\end{comment}

\LMnote{2010/09/28}{Added the \cs{CTANurl}'s and verified them}

%%% keep
\bibitem[Ars99]{URL}
  Donald Arseneau.
  \newblock \emph{The url package}.
  \newblock February, 1999.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/contrib/url/})

%%% keep
\bibitem[Ars01a]{TITLEREF}
  Donald Arseneau.
  \newblock \emph{The titleref package}.
  \newblock April, 2001.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/contrib/titleref/})

%%% keep
\bibitem[Ars01b]{CHAPTERBIB}
  Donald Arseneau.
  \newblock \emph{The chapterbib package}.
  \newblock September, 2001.
  \newblock (Available from CTAN via 
             \CTANurl{/macros/latex/contrib/cite/})

%%% keep
\bibitem[Ars07]{FRAMED}
  Donald Arseneau.
  \newblock \emph{The framed package} v0.95.
  \newblock October, 2007.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/contrib/framed/})

\begin{comment}
\bibitem[Bar92]{BAROLINI92}
  Helen Barolini.
  \newblock \emph{Aldus and his Dream Book}.
  \newblock Italica Press, 1992.
  \newblock ISBN 0--934977--22--4.

\bibitem[Bar01]{BARTRAM01}
  Alan Bartram.
  \newblock \emph{Five hundred years of book design}.
  \newblock Yale university Press, 2001.
  \newblock ISBN 0--300--09058--7.
  \newblock (First published 2001 by The British Library)

\bibitem[BDG89]{BIGELOW89}
  Charles Bigelow, Paul Hayden Duensing and Linnea Gentry (Eds).
  \newblock \emph{Fine Print on Type}. 1989.
  \newblock Fine Print, CA (ISBN 0--9607290-X) or
  \newblock Bedford Arts, CA (ISBN 0--938491--17--2).

\bibitem[Boh90]{BOHLE90}
  Robert Bohle.
  \newblock \emph{Publication Design for Editors}.
  \newblock Prentice-Hall,
  \newblock 1990.
\end{comment}

%%% keep
\bibitem[Ber02]{JURABIB}
  Jens Berger.
  \newblock \emph{The titlesec and titletoc packages}.
  \newblock September, 2002.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/contrib/titlesec/})

%%% keep
\bibitem[Bez99]{TITLESEC}
  Javier Bezos.
  \newblock \emph{The titlesec and titletoc packages}.
  \newblock February, 1999.
  \newblock (Available from CTAN via 
             \CTANurl{/macros/latex/contrib/titlesec/})

\bibitem[Bir04]{BIRDSALL04}
  Derek Birdsall.
  \newblock \emph{notes on book design}.
  \newblock Yale University Press, 2004.
  \newblock ISBN 0--300--10347--6.

\begin{comment}
\bibitem[Bra94]{MAKEIDX}
  Johannes Braams \textit{et al}.
  \newblock \emph{Standard LaTeX2e packages makeidx and showidx}.
  \newblock November, 1994.
  \newblock (Available from CTAN as 
             \url{/macros/latex/base/makeindx.dtx(ins)})
\end{comment}

%%% keep
\bibitem[Bra97]{ALLTT}
  Johannes Braams.
  \newblock \emph{The alltt environment}.
  \newblock June, 1997.
  \newblock (Available as \texttt{alltt.dtx} and 
  \texttt{alltt.ins} from CTAN via  \CTANurl{/macros/latex/base/})

%%% keep
\bibitem[Bri99]{BRINGHURST99}
  Robert Bringhurst.
  \newblock \emph{The Elements of Typographic Style}.
  \newblock Hartley \& Marks, second edition,
  \newblock 1999. ISBN 0--88179--033--8.

\begin{comment}
\bibitem[Bur59]{BURT59}
  C.~L.~Burt.
  \newblock \emph{A Psychological Study of Typography}.
  \newblock Cambridge University Press,
  \newblock 1959.
\end{comment}

%%% keep
\bibitem[Car14]{DELARRAY}
  David Carlisle.
  \newblock \emph{The delarray package}.
  \newblock October, 2014.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/required/tools/})

%%% keep
\bibitem[Car95]{AFTERPAGE}
  David Carlisle.
  \newblock \emph{The afterpage package}.
  \newblock October, 1995.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/required/tools/})

%%% keep
\bibitem[Car98b]{LONGTABLE}
  David Carlisle.
  \newblock \emph{The longtable package}.
  \newblock May, 1998.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/required/tools/})

%%% keep
\bibitem[Car98c]{ENUMERATE}
  David Carlisle.
  \newblock \emph{The enumerate package}.
  \newblock August, 1998.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/required/tools/})

\begin{comment}
\bibitem[Car98d]{REMRESET}
  David Carlisle.
  \newblock \emph{The remreset package}.
  \newblock August, 1998.
  \newblock (Available from CTAN in
             \url{/macros/latex/contrib/carlisle})
\end{comment}

%%% keep
\LMnote{2018/09/06}{Updated year}
\bibitem[Car16]{TABULARX}
  David Carlisle.
  \newblock \emph{The tabularx package}.
  \newblock February, 2016.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/required/tools/})

%%% keep
\bibitem[CR99]{GRAPHICX}
  David Carlisle and Sebastian Rahtz.
  \newblock \emph{The graphicx package}.
  \newblock February, 1999.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/required/graphics/})

%%% keep
\LMnote{2018/09/06}{updated year}             
\bibitem[Car14]{DCOLUMN}
  David Carlisle.
  \newblock \emph{The dcolumn package}.
  \newblock May, 2001.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/required/tools/})

\bibitem[Car04]{textcase} % casing intentional
 David Carlisle.
\newblock \emph{The \Lpack{textcase} package}.
\newblock October, 2004.
\newblock (Available from CTAN in 
      \texttt{/macros/latex/contrib/textcase})



%%% keep
\bibitem[Car05]{COLOR}
  David Carlisle.
  \newblock \emph{Packages in the graphics bundle} (includes the color package).
  \newblock November, 2005.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/required/graphics/})

\begin{comment}
%%% keep, not anymore, the bidi package is a bit different now, Vafa
%%% is the maintainer
             \bibitem[CK09]{BIDI} Fran\c{c}ois Charette and Vafa
               Khalighi.  \newblock \emph{Bidi: A convenient interface
                 for typesetting bidirectional texts with XeLaTeX}.
               \newblock 2009.  \newblock (Available from CTAN via
               \CTANurl{/macros/latex/contrib/bidi/})
\end{comment}




%%% keep
\bibitem[CB99]{CHAPPELL99}
  Warren Chappell and Robert Bringhurst.
  \newblock \emph{A Short History of the Printed Word}.
  \newblock Hartley \& Marks, 1999.
  \newblock ISBN 0--88179--154--7.

%%% keep
\bibitem[CH88]{CHEN88}
  Pehong Chen and Michael A.~Harrison.
  \newblock `Index Preparation and Processing'.
  \newblock \emph{Software: Practice and Experience}, 19:8, pp. 897--915,
            September, 1988.
  \newblock (Available from CTAN via
             \CTANurl{/indexing/makeindex/paper/})

\begin{comment}
\bibitem[Che05]{CHENG05}
  Karen Cheng.
  \newblock \emph{Designing Type}.
  \newblock Yale University Press, 2005.
  \newblock ISBN 0--300--11150--9.
\end{comment}

%%% keep
\bibitem[Chi93]{CMS}
  \newblock \emph{The Chicago Manual of Style}, Fourteenth Edition.
  \newblock The University of Chicago, 1993.
  \newblock ISBN 0--226--10389--7.


\begin{comment}
\bibitem[Coc02]{SUBFIGURE}
  Steven Douglas Cochran.
  \newblock \emph{The subfigure package}.
  \newblock March, 2002.
  \newblock (Available from CTAN in
             \url{/macros/latex/contrib/subfigure})

\bibitem[CG96]{CONWAY96}
  John H.~Conway and Richard K.~Guy.
  \newblock \emph{The Book of Numbers}.
  \newblock Copernicus, Springer-Verlag, 1996.
  \newblock ISBN 0--387--97993--X.

\bibitem[Cra92]{CRAIG92}
  James Craig.
  \newblock \emph{Designing with Type: A Basic Course in Typography}.
  \newblock Watson-Guptill, NY,
  \newblock 1992.
\end{comment}

%%% keep
\bibitem[Dal99a]{NATBIB}
  Patrick W.~Daly.
  \newblock \emph{Natural Sciences Citations and References}.
  \newblock May, 1999.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/contrib/natbib/})

%%% keep
\bibitem[Dal99b]{MAKEBST}
  Patrick W.~Daly.
  \newblock \emph{Customizing Bibliographic Style Files}.
  \newblock August, 1999.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/contrib/custom-bib})

\begin{comment}
\bibitem[Deg92]{DEGANI92}
  Asaf Degani.
  \newblock \emph{On the Typography of Flight-Deck Documentation}.
  \newblock NASA Contractor Report \# 177605.
  \newblock December, 1992.
  \newblock (Available from 
             \url{http://members.aol.com/willadams/typgrphy.htm#NASA})
\end{comment}

%%% keep
\bibitem[Dow96]{DOWDING96}
  Geoffrey Dowding.
  \newblock \emph{Finer Points in the Spacing \& Arrangement of Type}.
  \newblock Hartley \& Marks, 1996.
  \newblock ISBN 0--88179--119--9.

\begin{comment}
\bibitem[Dow98]{DOWDING98}
  Geoffrey Dowding.
  \newblock \emph{An Introduction to the History of Printing Types}.
  \newblock The British Library and Oak Knoll Press, 1998.
  \newblock ISBN 0--7123--4563--9 \textsc{uk},
                  1--884718--44--2 \textsc{usa}.
\end{comment}

%%% keep
\bibitem[Dow00]{PATCHCMD}
  Michael J.~Downes.
  \newblock \emph{The patchcmd package}.
  \newblock July, 2000.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/contrib/patchcmd/})

%%% keep
\bibitem[Eij92]{TEXBYTOPIC}
  Victor Eijkhout.
  \newblock \emph{TeX by Topic}.
  \newblock Addison-Wesley, 1992.
  \newblock ISBN 0--201--56882--9.
  \newblock (Available from \url{http://www.eijkhout.net/tbt/}).

%%% keep
\bibitem[Eij99]{COMMENT}
  Victor Eijkhout.
  \newblock \emph{comment.sty}
  \newblock October, 1999.
  \newblock (Available from CTAN via
            \CTANurl{/macros/latex/contrib/comment/})

\begin{comment}
\bibitem[Fai98]{MOREVERB}
  Robin Fairbairns.
  \newblock \emph{The moreverb package}.
  \newblock December, 1998.
  \newblock (Available from CTAN in
            \url{/macros/latex/contrib/moreverb})
\end{comment}

%%% keep
\bibitem[Fai00]{FOOTMISC}
  Robin Fairbairns.
  \newblock \emph{footmisc --- a portmanteau package for customising
                  footnotes in LaTeX2e}.
  \newblock March, 2000.
  \newblock (Available from CTAN via
            \CTANurl{/macros/latex/contrib/footmisc/})

%%% keep
\LMnote{2010/09/28}{link changed}
\bibitem[FAQ]{FAQ}
  Robin Fairbairns.
  \newblock \emph{The UK TeX FAQ}.
  \newblock (Available from CTAN via
            \url{http://faq.tug.org/})

\LMnote{2010/02/08}{Added from the main bib file}
\bibitem[Far90]{FARIN90}
Gerald Farin.
\newblock \emph{Curves and Surfaces for Computer Aided Geometric
  Design --- A Practical Guide}.
\newblock Academic Press, 2 edition, 1990.

\LMnote{2010/02/08}{Added from the main bib file}
\bibitem[FP80]{FAUX80}
I.~D. Faux and M.~J. Pratt.
\newblock {\em Computational Geometry for Design and Manufacture}.
\newblock Ellis Horwood, 1980.



%%% keep
\LMnote{2018/09/06}{Updated year}
\bibitem[Fea16]{BOOKTABS}
  Simon Fear.
  \newblock \emph{Publication quality tables in LaTeX}.
  \newblock April, 2016.
  \newblock (Available from CTAN via
            \CTANurl{/macros/latex/contrib/booktabs})

\begin{comment}
\bibitem[Fli98]{LETTRINE}
  Daniel Flipo.
  \newblock \emph{Typesetting `lettrines' in LaTeX2e documents}.
  \newblock March, 1998.
  \newblock (Available from CTAN in
             \url{/macros/latex/contrib/lettrine})
\end{comment}

%%% keep
\bibitem[Fly98]{FLYNN02}
  Peter Flynn.
  \newblock \emph{Formatting Information: A Beginner's Introduction to 
                  Typesetting with LaTeX2}.
  \newblock 2002.
  \newblock (Available from CTAN via 
             \CTANurl{/info/beginlatex/})

\begin{comment}
\bibitem[Fra00]{CROP}
  Melchior Franz.
  \newblock \emph{The crop package}.
  \newblock February, 2000.
  \newblock (Available from CTAN in 
             \url{/macros/latex/contrib/crop})

\bibitem[FOS98]{FRIEDL98}
  Friedrich Friedl, Nicolaus Ott and Bernard Stein.
  \newblock \emph{Typography: An Encyclopedic Survey of Type Designs and
                  Techniques throughout History}.
  \newblock Black Dog \& Leventhal Publishers Inc., 1998.
  \newblock ISBN 1--57912--023--7.

\bibitem[Gar66]{GARDNER66}
  Martin Gardner.
  \newblock \emph{More Mathematical Puzzles and Diversions}.
  \newblock Penguin Books, 1996.
  \newblock ISBN 0--14--020748--1.
\end{comment}

%\bibitem[GMS94]{GOOSSENS94}
%  Michel Goossens, Frank Mittelbach and Alexander Samarin.
%  \newblock \emph{The LaTeX Companion}.
%  \newblock Addison-Wesley Publishing Company, 1994
%  \newblock (ISBN 0--201--54199--8), 1994.

%%% keep
\bibitem[GM\textsuperscript{+}07]{GCOMPANION}
  Michel Goossens, Frank Mittelbach, et al.
  \newblock \emph{The LaTeX Graphics Companion: Second edition}.
  \newblock Addison-Wesley, 2007.
  \newblock ISBN 0--321--50892--0.

%%% keep
\bibitem[GR99]{WCOMPANION}
  Michel Goossens and Sebastian Rahtz (with Eitan Gurari,
  Ross Moore and Robert Sutor).
  \newblock \emph{The LaTeX Web Companion: Integrating TeX, HTML and XML}.
  \newblock Addison-Wesley, 1999.
  \newblock ISBN 0--201--43311--7.

\begin{comment}
\bibitem[Gou87]{GOULD87}
  J.~D.~Gould \textit{et al}.
  \newblock `Reading from CRT displays can be as fast as reading from paper'.
  \newblock \emph{Human Factors}, pp 497--517, 29:5, 1987.

\bibitem[HR83]{HARTLEY83}
  J.~Hartley and D.~Rooum.
  \newblock `Sir Cyril Burt and typography'.
  \newblock \emph{British Journal of Psychology}, pp 203--212, 74:2, 1983.

\bibitem[HM01]{HELLER01}
  Steven Heller and Philip B.~Meggs (Eds).
  \newblock \emph{Texts on Type: Critical Writings on Typography}.
  \newblock Allworth Press, 2001.
  \newblock ISBN 1--58115--082--2.
\end{comment}

%%% keep
\bibitem[Hoe98]{HOENIG98}
  Alan Hoenig.
  \newblock \emph{TeX Unbound: LaTeX and TeX strategies for fonts,
                  graphics, and more}.
  \newblock Oxford University Press, 1998.
  \newblock ISBN 0--19--509686--X.

\begin{comment}
\bibitem[HK75]{HVISTENDAHL75}
  J.~K.~Hvistendahl and M.~R.~Kahl.
  \newblock `Roman vs. sans serif body type: Readability and reader prference'.
  \newblock \emph{AANPA News Research Bulletin}, pp 3--11, 17 Jan., 1975.
\end{comment}

%%% keep
\bibitem[Jon95]{INDEX}
  David M.~Jones.
  \newblock \emph{A new implementation of LaTeX's indexing commands}.
  \newblock September, 1995.
  \newblock (Available from CTAN via 
  \CTANurl{/macros/latex/contrib/camel})

%%% keep
\bibitem[Keh98]{XINDY}
  Roger Kehr.
  \newblock \emph{xindy: A flexible indexing system}.
  \newblock February, 1998.
  \newblock (Available from CTAN via \CTANurl{/indexing/xindy/})

%%% keep
\bibitem[Ker07]{XCOLOR}
  Uwe Kern.
  \newblock \emph{Extending LaTeX's color facilities: the xcolor package}.
  \newblock January, 2007.
  \newblock (Available from CTAN via 
             \CTANurl{/macros/latex/contrib/xcolor/})

%%% keep, moved herer /daleif 2020/09/28
\bibitem[Kha10]{BIDI} Vafa
  Khalighi.  
  \newblock \emph{The \textsf{Bidi} package}.
  \newblock 2010.  \newblock (Available from CTAN via
  \CTANurl{/macros/latex/contrib/bidi/})



%%% keep
\bibitem[Knu84]{TEXBOOK}
  Donald E.~Knuth.
  \newblock \emph{The TeXbook}.
  \newblock Addison-Wesley, 1984.
  \newblock ISBN 0--201--13448--9.

%%% keep
\bibitem[Knu86]{TEXPROGRAM}
  Donald E.~Knuth.
  \newblock \emph{TeX: The Program}.
  \newblock Addison-Wesley, 1986.
  \newblock ISBN 0--201--13437--3.

%%% keep
\bibitem[Knu87]{CM}
  Donald E.~Knuth.
  \newblock \emph{Computer Modern Typefaces}.
  \newblock Addison-Wesley, 1987.
  \newblock ISBN 0--201--134446--2.

%%% keep
\bibitem[Knu92]{METAFONT}
  Donald E.~Knuth.
  \newblock \emph{The METAFONT Book}.
  \newblock Addison-Wesley, 1992.
  \newblock ISBN 0--201--13444--6.

%%% keep
\bibitem[Lam94]{LAMPORT94}
  Leslie Lamport.
  \newblock \emph{LaTeX: A Document Preparation System}.
  \newblock Addison-Wesley, 1994.
  \newblock ISBN 0--201--52983--1.

%%% keep
\bibitem[LEB04]{NTG}
  Leslie Lamport, Victor Eijkhout and Johannes Braams.
  \newblock \emph{NTG document classes for LaTeX version 2e}.
  \newblock June, 2004.
  \newblock (Available from CTAN via
  \CTANurl{/macros/latex/contrib/ntgclass/}) 

\begin{comment}
\bibitem[LMB99]{CLASSES}
  Leslie Lamport, Frank Mittelbach and Johannes Braams.
  \newblock \emph{Standard document classes for LaTeX version 2e}.
  \newblock September, 1999.
  \newblock (Available from CTAN as \url{/macros/latex/base/classes.dtx})

\bibitem[Law90]{LAWSON90}
  Alexander Lawson.
  \newblock \emph{Anatomy of a Typeface}.
  \newblock David R.~Godine, 1990. 
  \newblock ISBN 0--87923--333--8.

\bibitem[LA90]{LAWSONAGNER90}
  Alexander S.~Lawson with Dwight Agner.
  \newblock \emph{Printing Types: An Introduction}.
  \newblock Beacon Press, 1990. 
  \newblock ISBN 0--8070--6661--3.
\end{comment}

%%% keep
\bibitem[Lea03]{TEXFONTSITE}
  Christopher League.
  \newblock \emph{TeX support for the FontSite 500 CD}.
  \newblock May 2003.
  \newblock (Available from 
             \url{http://contrapunctus.net/fs500tex})

%%% keep
\bibitem[Leh04]{FONTINST}
  Philipp Lehman.
  \newblock \emph{The Font Installation Guide}.
  \newblock December 2004.
  \newblock (Available from CTAN via
             \CTANurl{/info/Type1fonts/fontinstallationguide})

%%% keep
\bibitem[Leu92]{LEUNEN92}
  Mary-Claire van Leunen.
  \newblock \emph{A Handbook for Scholars}.
  \newblock Oxford University Press, 1992.
  \newblock ISBN 0--19--506954--4.

\begin{comment}
\bibitem[Liv02]{LIVIO02}
  Mario Livio.
  \newblock \emph{The Golden Ratio: The Story of Phi, the World's Most
     Astonishing Number}.
  \newblock Broadway Books, 2002.
  \newblock ISBN 0--7679--0816--3.
\end{comment}

%%% keep
\bibitem[Lon91]{MULTIND}
  F.~W.~Long.
  \newblock \emph{multind}.
  \newblock August, 1991.
  \newblock (Available from CTAN as
  \CTANurl{/macros/latex209/contrib/misc/multind.sty}) 

%%% keep
\bibitem[Mad06]{CHAPSTYLES}
  Lars Madsen.
  \newblock \emph{Various chapter styles for the memoir class}.
  \newblock July, 2006.
  \newblock (Available from CTAN via
  \CTANurl{/info/latex-samples/MemoirChapStyles/}) 

%%% keep
\bibitem[Mad07]{MEMEXSUPP}
  Lars Madsen.
  \newblock \emph{The Memoir Experimental Support Package}.
  \newblock 2007.
  \newblock (Available from CTAN via
  \CTANurl{/macros/latex/contrib/memexsupp/}) 

%%% keep
\bibitem[McD98]{SECTSTY}
  Rowland McDonnell.
  \newblock \emph{The sectsty package}.
  \newblock November, 1998.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/contrib/sectsty/})

\begin{comment}
\bibitem[McL75]{MCLEAN75}
  Ruari McLean.
  \newblock \emph{Jan Tschichold: Typographer}.
  \newblock David R.~Godine, 1975.
  \newblock ISBN 0--87923--841--0.
\end{comment}

%%% keep
\bibitem[McL80]{MCLEAN80}
  Ruari McLean.
  \newblock \emph{The Thames \& Hudson Manual of Typography}.
  \newblock Thames \& Hudson, 1980.
  \newblock ISBN 0--500--68022--1.

\begin{comment}
\bibitem[McL95]{MCLEAN95}
  Ruari McLean (Ed).
  \newblock \emph{Typographers on Type}.
  \newblock W.~W.~Norton \& Co., 1995.
  \newblock ISBN 0--393--70201--4.

\bibitem[MMc95]{MEGGS00}
  Philip B.~Meggs and Roy McKelvey (Eds).
  \newblock \emph{Revival of the Fittest: Digital Versions of the Classic Typefaces}.
  \newblock RC Publications, Inc., 2000.
  \newblock ISBN 1--883915--08--2.

\bibitem[Mit95]{DOCSHORTVRB}
  Frank Mittelbach.
  \newblock \emph{The doc and shortvrb packages}.
  \newblock May, 1995.
  \newblock (Available from CTAN in 
            \url{/macros/latex/base})
\end{comment}

%%% keep
\bibitem[Mit18]{MULTICOL}
  Frank Mittelbach.
  \newblock \emph{An environment for multicolumn output}.
  \newblock April, 2018.
  \newblock (Available from CTAN (the \texttt{multicol} package) via 
            \CTANurl{/macros/latex/required/tools/})

%%keep
\LMnote{2018/09/06}{Updated year}            
\bibitem[MC18]{ARRAY}
  Frank Mittelbach and David Carlisle.
  \newblock \emph{A new implementation of LaTeX's tabular and array environment}.  \newblock May, 2018.
  \newblock (Available from CTAN (the \texttt{array} package) via
            \CTANurl{/macros/latex/required/tools/})

%%% keep
\bibitem[MC00]{FIXLTX2E}
  Frank Mittelbach and David Carlisle.
  \newblock \emph{The fixltx2e package}.
  \newblock December, 2016.
  \newblock As of 2015 the functionality of this package has been
  merged with the \LaTeX{} kernel.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/base/})

%%% keep
\bibitem[MG\textsuperscript{+}04]{COMPANION}
  Frank Mittelbach, Michael Goossens, et al.
  \newblock \emph{The LaTeX Companion: Second Edition}.
  \newblock Addison-Wesley, 2004.
  \newblock ISBN 0--201--36299--6.

\begin{comment}
\bibitem[Mor99]{MORISON99}
  Stanley Morison.
  \newblock \emph{A Tally of Types}.
  \newblock David R. Godine, 1999.
  \newblock ISBN 1--56792--004--7.
\end{comment}

\LMnote{2010/02/08}{Added from the main bib file}
\bibitem[Mor85]{MORTENSON85}
Michael~E. Mortenson.
\newblock \emph{Geometric Modeling}.
\newblock John Wiley \& Sons, 1985.


%%% keep
\bibitem[NG98]{SIDECAP}
  Rolf Niespraschk and Hubert G\"{a}\ss{}lein. 
  \newblock \emph{The sidecap package}.
  \newblock June, 1998.
  \newblock (Available from CTAN via
            \CTANurl{/macros/latex/contrib/sidecap/})

%%% keep
\bibitem[Oet]{LSHORT}
  Tobias Oetiker.
  \newblock \emph{The Not So Short Introduction to LaTeX2e}.
  \newblock (Available from CTAN via 
            \CTANurl{/info/lshort/})

%%% keep
\bibitem[Oos96]{FANCYHDR}
  Piet van Oostrum.
  \newblock \emph{Page Layout in LaTeX}.
  \newblock June, 1996.
  \newblock (Available from CTAN via
            \CTANurl{/macros/latex/contrib/fancyhdr/})

%%% keep
\bibitem[Pak01]{SYMBOLS}
  Scott Pakin.
  \newblock \emph{The Comprehensive LaTeX Symbol List}.
  \newblock July, 2001.
  \newblock (Available from CTAN via
            \CTANurl{/info/symbols/comprehensive/})

%%% keep
\bibitem[dP84]{PARVILLE84}
  H.~de~Parville.
  \newblock Recreations mathematique: {La Tour d'Hanoi} et la question 
            du {Tonkin}.
  \newblock \emph{La Nature}, part {I}:285--286, Paris 1884.

%%% keep
\bibitem[Pat88a]{BIBTEX}
  Oren Patashnik.
  \newblock \emph{BibTeXing}.
  \newblock February, 1988.
  \newblock (Available from CTAN as 
            \CTANurl{/bibliography/bibtex/distribs/doc/btxdoc.tex})

%%% keep
\bibitem[Pat88b]{BIBTEXHACK}
  Oren Patashnik.
  \newblock \emph{Designing BibTeX Styles}.
  \newblock February, 1988.
  \newblock (Available from CTAN as 
            \CTANurl{/bibliography/bibtex/distribs/doc/btxhak.tex})

%%% keep
\bibitem[Pug02]{MATHPAZO}
  Diego Puga.
  \newblock \emph{The Pazo Math fonts for mathematical typesetting
                  with the Palatino fonts}.
  \newblock May, 2002.
  \newblock (Available from CTAN via
            \CTANurl{/fonts/mathpazo/})

%%% keep
\bibitem[Rahtz01]{NAMEREF}
  Sebastian Rahtz.
  \newblock \emph{Section name references in LaTeX}.
  \newblock January, 2001. 
  \newblock (Available from CTAN (the \texttt{nameref} package) via 
            \CTANurl{/macros/latex/contrib/hyperref/})

%%% keep
\bibitem[Rahtz02]{HYPERREF}
  Sebastian Rahtz.
  \newblock \emph{Hypertext marks in LaTeX}.
  \newblock May, 2002. Now maintained and frequently updated by Heiko Oberdiek.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/hyperref/})

%%% keep
\bibitem[Rec97]{EPSLATEX}
  Keith Reckdahl.
  \newblock \emph{Using Imported Graphics in LaTeX2e}.
  \newblock December, 1997. Updated in 2006.
  \newblock (Available from CTAN via
              \CTANurl{/info/epslatex.pdf})

\begin{comment}
\bibitem[Reh72]{REHE72}
  Rolf Rehe.
  \newblock `Type and how to make it most legible'.
  \newblock \emph{Design Research International}, 1972.
\end{comment}

%%% keep
\bibitem[Rei07]{REINGOLD07}
  Edward M. Reingold.
  \newblock `Writing numbers in words in TeX'.
  \newblock TUGboat, 28, 2 pp 256--259, 2007.

\begin{comment}
\bibitem[RAE71]{ROBINSON71}
  D.~O.~Robinson, M.~Abbamonte and S.~H.~Evans.
  \newblock `Why serifs are important: The perception of small print'.
  \newblock \emph{Visible Language}, pp 353--359, 4, 1971.

\bibitem[Rog43]{ROGERS43}
  Bruce Rogers.
  \newblock \emph{Paragraphs on Printing}.
  \newblock William E. Rudge's Sons, Inc., 1943.
  \newblock (Reissued by Dover, 1979, ISBN 0--486--23817--2)

\bibitem[Rog49]{ROGERS49}
  Bruce Rogers.
  \newblock \emph{Centaur Types}.
  \newblock October House, 1949.
\end{comment}

%%% keep
\bibitem[RBC74]{ROUSEBALL}
  W. W. Rouse Ball and H. S. M. Coxeter. 
  \newblock \emph{Mathematical Recreations and Essays}.
  \newblock University of Toronto Press, twelfth edition, 1974.

\begin{comment}
\bibitem[Ryd76]{RYDER}
  John Ryder. 
  \newblock \emph{Printing for Pleasure}. Revised edition.
  \newblock The Bodley Head, 1976.
  \newblock ISBN 0--370--10443--9.
  \newblock (In the USA published by Henry Regenery Co., Michigan, 1977.
             ISBN 0--8092--78103--3)
\end{comment}

%%% keep
\bibitem[SW94]{EBOOK}
  Douglas Schenck and Peter Wilson.
  \newblock \emph{Information Modeling the EXPRESS Way}.
  \newblock Oxford University Press, 1994.
  \newblock ISBN 0--19--508714--3.

%%% keep
\bibitem[SRR99]{VERBATIM}
  Rainer Sch\"{o}pf, Bernd Raichle and Chris Rowley.
  \newblock \emph{A New Implementation of LaTeX's verbatim
                  and verbatim* Environments}.
  \newblock December, 1999.
  \newblock (Available from CTAN via
            \CTANurl{/macros/latex/required/tools/})

\begin{comment}
\bibitem[Sch97]{SCHRIVER97}
  Karen A.~Schriver.
  \newblock \emph{Dynamics in Document Design}.
  \newblock Wiley \& Sons, 1997.

\bibitem[Sme96]{SMEIJERS96}
  Fred Smeijers.
  \newblock \emph{Counterpunch: making type in the sixteenth century,
                  designing typefaces now}.
  \newblock Hyphen Press, London, 1996.
  \newblock ISBN 0--907259--06--5.
\end{comment}


\bibitem[Sch07]{practex-2007-3-ms}
Martin Scharrer,
\newblock\emph{Version Control of LaTeX Documents with svn-multi}.
\newblock  \emph{The Prac{\TeX} Journal}, 3, 2007.
\newblock ISSN 1556-6994.

\bibitem[Sch09]{svn-multi}
Martin Scharrer, 
\newblock\emph{The \textsf{svn-multi} package}, 2009. 
\newblock (Available from CTAN via
\CTANurl{/macros/latex/contrib/svn-multi/}) 


%%% keep
\bibitem[Sne04]{SNEEP04}
  Maarten Sneep.
  \newblock \emph{The atmosphere in the laboratory: cavity ring-down
                  measurements on scattering and absorption}.
  \newblock Phd thesis,
  \newblock Vrijie Universiteit, Amsterdam, 2004. 

%%% keep
\bibitem[Tal06]{DATETIME}
  Nicola L. C. Talbot.
  \newblock \emph{datetime.sty: Formatting Current Date and Time}.
  \newblock December, 2006.
  \newblock (Available from CTAN via
            \CTANurl{/macros/latex/contrib/datetime/})

%%% keep
\bibitem[Thi98]{ORNAMENTAL}
  Christina Thiele.
  \newblock `Hey --- it works: Ornamental rules'. 
  \newblock \emph{TUGboat}, 
  \newblock vol. 19, no. 4, p 427, December 1998.

%%% keep
\bibitem[Thi99]{TTC199}
  Christina Thiele.
  \newblock `The Treasure Chest: Package tours from CTAN', 
  \newblock \emph{TUGboat}, 
  \newblock vol. 20, no. 1, pp 53--58, March 1999.

%%% keep
\bibitem[TJ05]{CALC}
  Kresten Krab Thorup, Frank Jensen (and Chris Rowley).
  \newblock \emph{The calc package --- Infix notation arithmetic in LaTeX}.
  \newblock August, 2005.
  \newblock (Available from CTAN via
            \CTANurl{/macros/latex/required/tools/})

\begin{comment}
\bibitem[Tin63]{TINKER63}
  Miles A.~Tinker.
  \newblock \emph{Legibility of Print}.
  \newblock Books on Demand (University Microfilms International), 1963.
\end{comment}

%%% keep
\bibitem[Tob00]{SETSPACE}
  Geoffrey Tobin.
  \newblock \emph{setspace.sty}.
  \newblock December, 2000.
  \newblock (Available from CTAN via
            \CTANurl{/macros/latex/contrib/setspace/})

%%% keep
\bibitem[Tsc91]{TSCHICHOLD91}
  Jan Tschichold.
  \newblock \emph{The Form of the Book}.
  \newblock Lund Humphries, 1991.
  \newblock ISBN 0--85331--623--6.

\begin{comment}
\bibitem[Tuf83]{TUFTE83}
  Edward R. Tufte.
  \newblock \emph{The Visual Display of Quantative Information}.
  \newblock Graphics Press, 1983.
\end{comment}

%%% keep
\bibitem[Ume99]{GEOMETRY}
  Hideo Umeki.
  \newblock \emph{The geometry package}.
  \newblock November, 1999.
  \newblock (Available from CTAN in
            \CTANurl{/macros/latex/contrib/geometry/})


\begin{comment}
\bibitem[Whe95]{WHEILDON95}
  Colin Wheildon.
  \newblock \emph{Type \& Layout}.
  \newblock Strathmore Press, 1995.
  \newblock ISBN 0--9624891--5--8.
\end{comment}

%%% keep
\bibitem[Wil00]{CATALOGUE}
  Graham Williams.
  \newblock \emph{The TeX Catalogue}.
  \newblock (Latest version on CTAN as 
  \CTANurl{/help/Catalogue/catalogue.html})

%%% keep
\bibitem[Wil93]{ADRIANWILSON93}
  Adrian Wilson.
  \newblock \emph{The Design of Books}.
  \newblock Chronicle Books, 1993.
  \newblock ISBN 0--8118--0304--X.


%%% keep
\bibitem[Wil99b]{TOCVSEC2}
  Peter Wilson.
  \newblock \emph{The tocvsec2 package}.
  \newblock January, 1999.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/tocvsec2/})

%%% keep
\bibitem[Wil00a]{EPIGRAPH}
  Peter Wilson.
  \newblock \emph{The epigraph package}.
  \newblock February, 2000.
  \newblock (Available from CTAN via
            \CTANurl{/macros/latex/contrib/epigraph/})

%%% keep
\bibitem[Wil00b]{ISOCLASS}
  Peter Wilson.
  \newblock \emph{LaTeX files for typesetting ISO standards}.
  \newblock February, 2000.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/isostds/iso/})

%%% keep
\bibitem[Wil00c]{NEXTPAGE}
  Peter Wilson.
  \newblock \emph{The nextpage package}.
  \newblock February, 2000.
  \newblock (Available from CTAN as 
            \CTANurl{/macros/latex/contrib/misc/nextpage.sty})

\begin{comment}
\bibitem[Wil00d]{NEEDSPACE}
  Peter Wilson.
  \newblock \emph{The needspace package}.
  \newblock March, 2000.
  \newblock (Available from CTAN as 
            \url{/macros/latex/contrib/misc/needspace.sty})
\end{comment}

%%% keep
\bibitem[Wil00e]{XTAB}
  Peter Wilson.
  \newblock \emph{The xtab package}.
  \newblock April 2000.
  \newblock (Available from CTAN via
             \CTANurl{/macros/latex/contrib/xtab})

%%% keep
\bibitem[Wil01a]{ABSTRACT}
  Peter Wilson.
  \newblock \emph{The abstract package}.
  \newblock February, 2001.
  \newblock (Available from CTAN via
            \CTANurl{/macros/latex/contrib/abstract/})

\begin{comment}
%%%  it no longer exist
\LMnote{2010/09/28}{Removed as it no longer exist}
\bibitem[Wil01b]{CHNGPAGE}
  Peter Wilson.
  \newblock \emph{The changepage package}.
  \newblock February, 2001.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/changepage/})

\bibitem[Wil01c]{APPENDIX}
  Peter Wilson.
  \newblock \emph{The appendix package}.
  \newblock March, 2001.
  \newblock (Available from CTAN in 
            \url{/macros/latex/contrib/appendix})
\end{comment}

%%% keep
\bibitem[Wil01d]{CCAPTION}
  Peter Wilson.
  \newblock \emph{The ccaption package}.
  \newblock March, 2001.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/ccaption/})

%%% keep
\bibitem[Wil01e]{CHNGCNTR}
  Peter Wilson.
  \newblock \emph{The chngcntr package}.
  \newblock April, 2001.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/chngcntr/})

%%% keep
\bibitem[Wil01f]{HANGING}
  Peter Wilson.
  \newblock \emph{The hanging package}.
  \newblock March, 2001.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/hanging/})

%%% keep
\bibitem[Wil01g]{TITLING}
  Peter Wilson.
  \newblock \emph{The titling package}.
  \newblock March, 2001.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/titling/})

%%% keep
\bibitem[Wil01h]{TOCBIBIND}
  Peter Wilson.
  \newblock \emph{The tocbibind package}.
  \newblock April, 2001.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/tocbibind/})

%%% keep
\bibitem[Wil01i]{TOCLOFT}
  Peter Wilson.
  \newblock \emph{The tocloft package}.
  \newblock April, 2001.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/tocloft/})


\begin{comment}
\bibitem[Wil01j]{MEMOIR}
  Peter Wilson.
  \newblock \emph{The LaTeX memoir class for configurable book 
                  typesetting: Source code}.
  \newblock July, 2001.
  \newblock (Available from CTAN in 
            \url{/macros/latex/contrib/memoir})

\bibitem[Wil01k]{VERSE}
  Peter Wilson.
  \newblock \emph{Typesetting simple verse with LaTeX}
  \newblock July, 2001.
  \newblock (Available from CTAN in 
            \url{/macros/latex/contrib/verse})

\bibitem[Wil01l]{BOOKLET}
  Peter Wilson.
  \newblock \emph{Printing booklets with LaTeX}
  \newblock August, 2001.
  \newblock (Available from CTAN in 
            \url{/macros/latex/contrib/booklet})
\end{comment}

%%% keep
\bibitem[Wil03a]{LAYOUTS}
  Peter Wilson.
  \newblock \emph{The layouts package}
  \newblock November, 2003.
  \newblock (Available from CTAN in 
            \CTANurl{/macros/latex/contrib/layouts/})

%%% keep
\bibitem[Wil03b]{LEDMAC}
  Peter Wilson.
  \newblock \emph{ledmac: A presumptuous attempt to port EDMAC and
    TABMAC to LaTeX}
  \newblock November, 2003.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/ledmac/})

\begin{comment}
\bibitem[Wil04a]{GLISTER3}
  Peter Wilson.
  \newblock `Glisterings'.
  \newblock TUGboat, 25, 2 pp 201--202, 2004.
\end{comment}


\LMnote{2010/02/08}{Added from the main bib file}
\bibitem[Wil04a]{BEZ123}
Peter Wilson.
\newblock \emph{The bez123 and multiply packages}, April 2004.
\newblock (Available from CTAN in  \CTANurl{/macros/latex/contrib/bez123/})


%%% keep
\bibitem[Wil04b]{PAGENOTE}
  Peter Wilson.
  \newblock \emph{The pagenote package}
  \newblock September, 2004.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/pagenote/})




%%% keep
\bibitem[Wil07a]{TITLEPAGES}
  Peter Wilson.
  \newblock \emph{Some Examples of Title Pages}.
  \newblock Herries Press, 2007.
  \newblock (Available from CTAN via
  \CTANurl{/info/latex-samples/TitlePages/}) 

%%% keep
\begin{comment}
    % repeated below
  \bibitem[Wil07b]{MEMCODE} Peter Wilson.  \newblock \emph{The \ltx\
      memoir class for configurable book typesetting: source code}
    \newblock November, 2007.  \newblock (Available from CTAN in
    \url{/macros/latex/contrib/memoir})
\end{comment}
%%% keep

%%% keep, outcommented, as we do not use an addendum at the moment
\LMnote{2010/09/28}{out commented as we do not provide one at the moment}
\begin{comment}
\bibitem[Wil07d]{MEMADD}
Peter Wilson.
  \newblock \emph{ADDENDUM: The Memoir Class for Configurable
    Typesetting --- User Guide} 
  \newblock November, 2007.
  \newblock (Available from CTAN in 
            \url{/macros/latex/contrib/memoir})
\end{comment}
          
%%% keep
\bibitem[Wil07e]{GLISTER07}
Peter Wilson.
  \newblock `Glisterings', \emph{TUGboat}, 28(2):229--232, 2007.

\begin{comment}
\bibitem[Wil07f]{TUGKEYNOTE07}
Peter Wilson.
  \newblock `Between then and now --- A meandering memoir', 
  \newblock \emph{TUGboat}, 28(3):280--298, 2007.
\end{comment}

%%% keep
\bibitem[Wil08a]{CHANGEPAGE}
  Peter Wilson.
  \newblock \emph{The changepage package}.
  \newblock March, 2008.
  \newblock (Available from CTAN via
            \CTANurl{/macros/latex/contrib/changepage/})

%%% keep
\bibitem[Wil08b]{GLISTER08}
Peter Wilson.
  \newblock `Glisterings', \emph{TUGboat}, 29(2):324--327, 2008.

%%% keep
\bibitem[Wil09a]{FONTTABLE}
  Peter Wilson.
  \newblock \emph{The fonttable package}
  \newblock April, 2009.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/fonttable/})

\bibitem[Wil09b]{MEMCODE}
Peter Wilson (with the assistance of Lars Madsen).
  \newblock \emph{The \ltx\ memoir class for configurable book
    typesetting: source code} 
  \newblock July, 2009.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/memoir/})

\bibitem[Wil07c]{MEMMAN}
Peter Wilson (with the assistance of Lars Madsen).
  \newblock \emph{The Memoir Class for Configurable Typesetting ---
    User Guide} 
  \newblock August, 2009. Regularly updated.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/memoir/})

\bibitem[Wil09d]{MEMDESIGN}
  Peter Wilson.
  \newblock \emph{A Few Notes on Book Design}
  \newblock August, 2009.
  \newblock (Available from CTAN via
            \CTANurl{/info/memdesign/})


%%% keep
\bibitem[Wil??]{RUMOUR}
Peter Wilson.
\newblock \emph{A Rumour of Humour: A scientist's commonplace book}.
\newblock To be published.


\bibitem[Wri18]{SIUNITX}
  Joseph Wright.
  \newblock \emph{Siunitx — A comprehensive (SI) units package}
  \newblock May, 2018.
  \newblock (Available from CTAN via \CTANurl{/macros/latex/contrib/siunitx})




\begin{comment}
\bibitem[Wis03]{WISHART03}
David Wishart.
\newblock \emph{The Printing of Mathematics}
\newblock in \emph{Type \& Typography: Highlights from \emph{Matrix}, the
       review for printers \& bibliophiles}.
\newblock Mark Batty Publisher, 2003.
\newblock ISBN 0--9715687--6--6.
\newblock (Originally published in \emph{Matrix 8}, 1988)

\bibitem[Wul53]{WULLING-FOOTNOTES}
  Emerson G. Wulling.
  \newblock \emph{A Comp's-Eye View of Footnotes}.
  \newblock Sumac Press, 1953.

\bibitem[Zac69]{ZACHRISSOM69}
  B.~Zachrissom.
  \newblock \emph{Studies in the Legibility of Printed Text}.
  \newblock Almqvist \& Wiksell, Stockholm, 1969.
\end{comment}

%%% keep
\bibitem[Zan98]{FANCYBOX}
  Timothy Van Zandt.
  \newblock \emph{Documentation for fancybox.sty: Box tips and tricks for 
             LaTeX}, 
  \newblock November, 1998.
  \newblock (Available from CTAN via 
            \CTANurl{/macros/latex/contrib/fancybox/})

\begin{comment}
\bibitem[Zap00]{ZAPF00}
  Hermann Zapf.
  \newblock \emph{The Fine Art of Letters}.
  \newblock The Grolier Club, 2000.
  \newblock ISBN 0--910672--35--0.
\end{comment}

\end{thebibliography}
